19Q1

### 2019-04-22 19Q1.Rmd

  # Merge mutation calls
  data <- merge(data, maf_df_filt, by = c("Gene", "Broad_ID"), all.x = TRUE)
  data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  ## Identify point mutations
  data$Keep_PtMut_Tests <- with(data, ifelse(Reference_Allele == 1 | is.na(Reference_Allele), TRUE, ifelse(Reference_Allele == 0 & Tumor_Seq_Allele1 == 1, TRUE, FALSE)))
  ## Add color columns for plotting
  if(database == "CCLE") {
    data$Color <- ifelse(data$Mutation_Status == 0, "palegreen3", "springgreen4")
    data$Color <- factor(data$Color)
  }
  if(database == "CTRP") {
    data$Color <- ifelse(data$Mutation_Status == 0, "slategray3", "steelblue4")
    data$Color <- factor(data$Color)
  }
  if(database == "GDSC") {
    data$Color <- ifelse(data$Mutation_Status == 0, "paleturquoise3", "turquoise4")
    data$Color <- factor(data$Color)
  }
  print("Merged mutation calls.")

@@

FilterForPointMuts <- function(dataframe, outfile) {
  # Filter for point mutations
  ptmuts <- filter(dataframe, Keep_PtMut_Tests == TRUE)
  # Remove columns that lead to duplicate mutation entries
  ptmuts <- subset(ptmuts, select = -c(Variant_Classification, Reference_Allele, Tumor_Seq_Allele1, Keep_PtMut_Tests))
  ptmuts$Gene <- ordered(ptmuts$Gene, levels = g2p_genes)
  ptmuts <- unique(ptmuts)
  saveRDS(ptmuts, outfile, compress = "xz")
  return(ptmuts)
}s
```

@@

Mutation calls (coding region, germline filtered):
```{r, eval=FALSE}
maf_raw <- read.delim("./data_munging/19Q1/depmap_19Q1_mutation_calls_20190305.csv.gz", header = TRUE, sep = ",", row.names = 1)
# Harmonize ID column name
colnames(maf_raw)[colnames(maf_raw) == "DepMap_ID"] <- "Broad_ID"
colnames(maf_raw)[colnames(maf_raw) == "Hugo_Symbol"] <- "Gene"
# Select columns
maf_df <- subset(maf_raw, select = c("Gene", "Broad_ID", "Variant_Classification", "Reference_Allele", "Tumor_Seq_Allele1"))
maf_df$Reference_Allele <- ifelse(maf_df$Reference_Allele == "-", 0, nchar(as.character(maf_df$Reference_Allele)))
maf_df$Tumor_Seq_Allele1 <- nchar(as.character(maf_df$Tumor_Seq_Allele1))
maf_df <- maf_df[maf_df$Variant_Classification != "Silent",]
# Add Mutation_Status column
maf_df$Mutation_Status <- 1

@@

ccle <- MergeOmicsG2P(database = "CCLE", outfile = "./data_munging/rds/ccle_data_19Q1_g2pgenes.rds")
ctrp <- MergeOmicsG2P(database = "CTRP", outfile = "./data_munging/rds/ctrp_data_19Q1_g2pgenes.rds")
gdsc <- MergeOmicsG2P(database = "GDSC", outfile = "./data_munging/rds/gdsc_data_19Q1_g2pgenes.rds")

ccle_ptmuts <- FilterForPointMuts(dataframe = ccle, outfile = "./data_munging/rds/ccle_data_19Q1_g2pgenes_ptmuts.rds")
ctrp_ptmuts <- FilterForPointMuts(dataframe = ctrp, outfile = "./data_munging/rds/ctrp_data_19Q1_g2pgenes_ptmuts.rds")
gdsc_ptmuts <- FilterForPointMuts(dataframe = gdsc, outfile = "./data_munging/rds/gdsc_data_19Q1_g2pgenes_ptmuts.rds")
```

@@

```{r}
ccle <- readRDS("./data_munging/rds/ccle_data_19Q1_g2pgenes.rds")
ctrp <- readRDS("./data_munging/rds/ctrp_data_19Q1_g2pgenes.rds")
gdsc <- readRDS("./data_munging/rds/gdsc_data_19Q1_g2pgenes.rds")
ccle_ptmuts <- readRDS("./data_munging/rds/ccle_data_19Q1_g2pgenes_ptmuts.rds")
ctrp_ptmuts <- readRDS("./data_munging/rds/ctrp_data_19Q1_g2pgenes_ptmuts.rds")
gdsc_ptmuts <- readRDS("./data_munging/rds/gdsc_data_19Q1_g2pgenes_ptmuts.rds")

@@

```{r, eval=FALSE}
crispr <- read.delim("./data_munging/19Q1/gene_effect_19Q1_20181115.csv.gz", sep = ",", header  = TRUE, check.names = FALSE)
# Remove Entrez gene IDs from colnames
colnames(crispr) <- gsub(" .*", "", colnames(crispr))
```

Merge annotation data:
```{r, eval=FALSE}
# Melt CRISPR dataset for merging
crispr_melt <- melt(crispr, id.vars = "DepMap_ID", measure.vars = colnames(crispr)[2:ncol(crispr)], variable.name = "Gene", value.name = "Score")
# Merge cell line metadata
crispr_melt <- merge(crispr_melt, ccl_meta, by = "DepMap_ID", all.x = TRUE)
crispr_melt <- merge(crispr_melt, crispr_meta, by = c("CCLE_Name", "DepMap_ID"), all.x = TRUE)

# Merge mutation annotations
maf_filt_crispr <- filter(maf_df, Gene %in% as.character(unique(crispr_melt$Gene)))
crispr_muts <- merge(crispr_melt, maf_filt_crispr, by = c("Gene", "CCLE_Name", "DepMap_ID"), all.x = TRUE)
crispr_muts$Gene <- factor(crispr_muts$Gene)
crispr_muts <- crispr_muts %>% mutate(Mutation_Status_Nonsilent = if_else(is.na(Mutation_Status_Nonsilent), "Wildtype", Mutation_Status_Nonsilent))
# Summarize number of mutant and Wildtype cell lines
crispr_muts_summ <- crispr_muts %>% group_by(Gene) %>%
  summarize(N_Nonsilent_Wildtype = sum(Mutation_Status_Nonsilent == "Wildtype"),
            N_Nonsilent_Mutant = sum(Mutation_Status_Nonsilent == "Mutant"))
# Merge test results back into full dataset, which restores information lost in the summarization
crispr_data <- merge(crispr_muts_summ, crispr_muts, by = "Gene")
# Add Color columns
crispr_data$Color_Nonsilent <- ifelse(crispr_data$Mutation_Status_Nonsilent == "Wildtype", "thistle", "darkorchid")
crispr_data$Color_Nonsilent <- factor(crispr_data$Color_Nonsilent)

# Cell line lineages
crispr_data <- merge(crispr_data, ccl_converter, by = c("CCLE_Name", "DepMap_ID"), all.x = TRUE)
levels(crispr_data$lineage_name) <- sort(levels(crispr_data$lineage_name), decreasing = TRUE)
# Copy number
crispr_data <- merge(crispr_data, cn_melt, by = c("Gene", "DepMap_ID"), all.x = TRUE)
# Gene expression (RPKM)
ge_crispr_18Q3 <- filter(ge_melt_18Q3, Gene %in% as.character(unique(crispr_melt$Gene)))
# ge_crispr_19Q1 <- filter(ge_melt_19Q1, Gene %in% as.character(unique(crispr_melt$Gene)))
crispr_data <- merge(crispr_data, ge_crispr_18Q3, by = c("Gene", "CCLE_Name", "DepMap_ID"), all.x = TRUE)
# crispr_data <- merge(crispr_data, ge_crispr_19Q1, by = c("Gene", "DepMap_ID"), all.x = TRUE)
saveRDS(crispr_data, "./../htscreens_giant_files/crispr_data_19Q1.rds", compress = "xz")

# Takes a really long time to run (4+ hours)
# crispr_data_ptmuts <- filter(crispr_data, Reference_Allele == 1 | Reference_Allele == 0 | is.na(Reference_Allele))
# crispr_data_ptmuts$Keep_PtMut_Tests <- ifelse(crispr_data_ptmuts$Reference_Allele == 1 | is.na(crispr_data_ptmuts$Reference_Allele), TRUE, ifelse(crispr_data_ptmuts$Tumor_Seq_Allele1 == 1, TRUE, FALSE))
# crispr_data_ptmuts <- filter(crispr_data_ptmuts, Keep_PtMut_Tests == TRUE)
# crispr_data_ptmuts <- subset(crispr_data_ptmuts, select = -c(Variant_Classification, Reference_Allele, Tumor_Seq_Allele1, Genome_Change, Keep_PtMut_Tests))
# crispr_data_ptmuts <- unique(crispr_data_ptmuts)
# saveRDS(crispr_data_ptmuts, "./data_munging/rds/crispr_data_19Q1_ptmuts.rds", compress = "xz")
```

```{r, eval=FALSE}
crispr_data <- readRDS("./../htscreens_giant_files/crispr_data_19Q1.rds")
# Run next two lines only if the file was made (see end of the last chunk)
# crispr_data_ptmuts <- readRDS("./../htscreens_giant_files/crispr_data_19Q1_ptmuts.rds")
# crispr_data_g2p <- filter(crispr_data_ptmuts, Gene %in% g2p_genes)

# Filter CRISPR for G2P genes in G2P
crispr_data_ptmuts <- filter(crispr_data, Gene %in% g2p_genes)
# Filter G2P CRISPR for point mutations
crispr_data_ptmuts <- filter(crispr_data_ptmuts, Reference_Allele == 1 | Reference_Allele == 0 | is.na(Reference_Allele))
crispr_data_ptmuts$Keep_PtMut_Tests <- ifelse(crispr_data_ptmuts$Reference_Allele == 1 | is.na(crispr_data_ptmuts$Reference_Allele), TRUE, ifelse(crispr_data_ptmuts$Tumor_Seq_Allele1 == 1, TRUE, FALSE))
crispr_data_ptmuts <- filter(crispr_data_ptmuts, Keep_PtMut_Tests == TRUE)
crispr_data_ptmuts <- subset(crispr_data_ptmuts, select = -c(Variant_Classification, Reference_Allele, Tumor_Seq_Allele1, Genome_Change, Keep_PtMut_Tests))
crispr_data_g2p <- unique(crispr_data_ptmuts)
crispr_data_g2p$Gene <- factor(crispr_data_g2p$Gene, levels = g2p_genes)
crispr_data_g2p$Mutation_Status_Nonsilent <- factor(crispr_data_g2p$Mutation_Status_Nonsilent, levels = c("Mutant", "Wildtype"))
```

>>>>

  # Merge mutation calls
  data <- merge(data, maf_df_filt, by = c("Gene", "Broad_ID"), all.x = TRUE)
  data$Mutation_Status <- ifelse(is.na(data$Mutation_Status), 0, data$Mutation_Status)
  ## Identify point mutations
  data$PointMutation <- ifelse(is.na(data$PointMutation), 0, data$PointMutation)
  ## Add color columns for plotting
  if(database == "CCLE") {
    data$Color <- ifelse(data$Mutation_Status == 0, "palegreen3", "springgreen4")
    data$Color <- factor(data$Color)
  }
  if(database == "CTRP") {
    data$Color <- ifelse(data$Mutation_Status == 0, "slategray3", "steelblue4")
    data$Color <- factor(data$Color)
  }
  if(database == "GDSC") {
    data$Color <- ifelse(data$Mutation_Status == 0, "paleturquoise3", "turquoise4")
    data$Color <- factor(data$Color)
  }
  print("Merged mutation calls.")

@@

FilterForPointMuts <- function(dataframe, outfile) {
  # Filter for point mutations
  ptmuts <- filter(dataframe, PointMutation == 1)
  # Remove columns that lead to duplicate mutation entries
  ptmuts <- subset(ptmuts, select = -c(Variant_Classification, Reference_Allele, Tumor_Seq_Allele1, PointMutation))
  ptmuts$Gene <- ordered(ptmuts$Gene, levels = g2p_genes)
  ptmuts <- unique(ptmuts)
  saveRDS(ptmuts, outfile, compress = "xz")
  return(ptmuts)
}
```

@@

Mutation calls (coding region, germline filtered):
```{r, eval=FALSE}
maf_raw <- read.delim("./data_munging/19Q1/depmap_19Q1_mutation_calls_20190305.csv.gz", header = TRUE, sep = ",", row.names = 1)
# Harmonize ID column name
colnames(maf_raw)[colnames(maf_raw) == "DepMap_ID"] <- "Broad_ID"
colnames(maf_raw)[colnames(maf_raw) == "Hugo_Symbol"] <- "Gene"
# Select columns
maf_df <- subset(maf_raw, select = c("Gene", "Broad_ID", "Variant_Classification", "Reference_Allele", "Tumor_Seq_Allele1"))
maf_df$Reference_Allele <- ifelse(maf_df$Reference_Allele == "-", 0, nchar(as.character(maf_df$Reference_Allele)))
maf_df$Tumor_Seq_Allele1 <- nchar(as.character(maf_df$Tumor_Seq_Allele1))
maf_df <- maf_df[maf_df$Variant_Classification != "Silent",]
# Add Mutation_Status column
maf_df$Mutation_Status <- 1
maf_df$PointMutation <- with(maf_df, ifelse(Reference_Allele == 1 | is.na(Reference_Allele), TRUE, ifelse(Reference_Allele == 0 & Tumor_Seq_Allele1 == 1, 1, 0)))

@@

# ccle <- MergeOmicsG2P(database = "CCLE", outfile = "./data_munging/rds/ccle_data_19Q1_g2pgenes.rds")
# ctrp <- MergeOmicsG2P(database = "CTRP", outfile = "./data_munging/rds/ctrp_data_19Q1_g2pgenes.rds")
# gdsc <- MergeOmicsG2P(database = "GDSC", outfile = "./data_munging/rds/gdsc_data_19Q1_g2pgenes.rds")

# ccle_ptmuts <- FilterForPointMuts(dataframe = ccle, outfile = "./data_munging/rds/ccle_data_19Q1_g2pgenes_ptmuts.rds")
# ctrp_ptmuts <- FilterForPointMuts(dataframe = ctrp, outfile = "./data_munging/rds/ctrp_data_19Q1_g2pgenes_ptmuts.rds")
gdsc_ptmuts <- FilterForPointMuts(dataframe = gdsc, outfile = "./data_munging/rds/gdsc_data_19Q1_g2pgenes_ptmuts.rds")
```

@@

```{r, eval=FALSE}
ccle <- readRDS("./data_munging/rds/ccle_data_19Q1_g2pgenes.rds")
ctrp <- readRDS("./../htscreens_giant_files/ctrp_data_19Q1_g2pgenes.rds")
gdsc <- readRDS("./data_munging/rds/gdsc_data_19Q1_g2pgenes.rds")
ccle_ptmuts <- readRDS("./data_munging/rds/ccle_data_19Q1_g2pgenes_ptmuts.rds")
ctrp_ptmuts <- readRDS("./../htscreens_giant_files/ctrp_data_19Q1_g2pgenes_ptmuts.rds")
gdsc_ptmuts <- readRDS("./data_munging/rds/gdsc_data_19Q1_g2pgenes_ptmuts.rds")

@@

```{r, eval=FALSE}
crispr <- read.delim("./data_munging/19Q1/gene_effect_corrected_20190305.csv.gz", sep = ",", header  = TRUE, check.names = FALSE)
# Remove Entrez gene IDs from colnames
colnames(crispr) <- gsub(" .*", "", colnames(crispr))
colnames(crispr)[1] <- "Broad_ID"
# Melt CRISPR dataset for merging
crispr_melt <- melt(crispr, id.vars = "Broad_ID", measure.vars = colnames(crispr)[2:ncol(crispr)], variable.name = "Gene", value.name = "Score")
# Merge cell line metadata
crispr_melt <- merge(crispr_melt, ccl_meta, by = "Broad_ID", all.x = TRUE)
# Merge mutation annotations
maf_filt_crispr <- filter(maf_df, Gene %in% as.character(unique(crispr_melt$Gene)))
maf_filt_crispr <- subset(maf_filt_crispr, select = -c(Variant_Classification, Reference_Allele, Tumor_Seq_Allele1))
maf_filt_crispr <- unique(maf_filt_crispr)
crispr_muts <- merge(crispr_melt, maf_filt_crispr, by = c("Gene","Broad_ID"), all.x = TRUE)
crispr_muts$Gene <- factor(crispr_muts$Gene)
crispr_muts$Mutation_Status <- ifelse(is.na(crispr_muts$Mutation_Status), 0, crispr_muts$Mutation_Status)
## Identify point mutations and wildtypes for testing
crispr_muts$PointMutation <- ifelse(is.na(crispr_muts$PointMutation), 1, crispr_muts$PointMutation)
# Summarize number of mutant and Wildtype cell lines
crispr_muts_summ <- crispr_muts %>% group_by(Gene) %>%
  summarize(N_Wildtype = length(Mutation_Status) - sum(Mutation_Status == 1),
            N_Mutant = sum(Mutation_Status == 1))
# Merge test results back into full dataset, which restores information lost in the summarization
crispr_data <- merge(crispr_muts_summ, crispr_muts, by = "Gene")
# Add Color columns
crispr_data$Color <- ifelse(crispr_data$Mutation_Status == 0, "thistle", "darkorchid")
crispr_data$Color <- factor(crispr_data$Color)

# Cell line lineages
# crispr_data <- merge(crispr_data, ccl_converter, by = c("CCLE_Name", "Broad_ID"), all.x = TRUE)
# levels(crispr_data$lineage_name) <- sort(levels(crispr_data$lineage_name), decreasing = TRUE)
# Copy number
crispr_data <- merge(crispr_data, cn_melt, by = c("Gene", "Broad_ID"), all.x = TRUE)
# Gene expression (RPKM)
ge_crispr <- filter(ge_melt, Gene %in% as.character(unique(crispr_melt$Gene)))
# ge_crispr_19Q1 <- filter(ge_melt_19Q1, Gene %in% as.character(unique(crispr_melt$Gene)))
crispr_data <- merge(crispr_data, ge_crispr, by = c("Gene", "Broad_ID"), all.x = TRUE)
# crispr_data <- merge(crispr_data, ge_crispr_19Q1, by = c("Gene", "Broad_ID"), all.x = TRUE)
saveRDS(crispr_data, "./../htscreens_giant_files/crispr_data_19Q1.rds", compress = "xz")
```

```{r, eval=TRUE}
crispr_data <- readRDS("./../htscreens_giant_files/crispr_data_19Q1.rds")
# Run next two lines only if the file was made (see end of the last chunk)
# crispr_data_ptmuts <- readRDS("./../htscreens_giant_files/crispr_data_19Q1_ptmuts.rds")

# Filter G2P CRISPR for point mutations
crispr_data_ptmuts <- filter(crispr_data, PointMutation == 1)
crispr_data_ptmuts <- subset(crispr_data_ptmuts, select = -c(PointMutation))
# Filter CRISPR for G2P genes in G2P
crispr_data_g2p <- filter(crispr_data_ptmuts, Gene %in% g2p_genes)
crispr_data_g2p <- unique(crispr_data_g2p)
crispr_data_g2p$Gene <- factor(crispr_data_g2p$Gene, levels = g2p_genes)
crispr_data_g2p$Mutation_Status <- factor(crispr_data_g2p$Mutation_Status)
```
